---
title: "Lab 1 - Redwood Data"
author: "Tiffany Tang"
date: "`r Sys.Date()`"
output: vthemes::vmodern
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

for (fname in list.files(here::here("R"), pattern = "*.R")) {
  source(here::here(file.path("R", fname)))
}

DATA_PATH <- here::here("data")
```

# Load Data {.tabset .tabset-vmodern}

```{r load-data}
# load dates data
dates_orig <- load_dates_data(DATA_PATH)

# load motes data
motes_orig <- load_mote_location_data(DATA_PATH)

# load redwood data
redwood_all_orig <- load_redwood_data(DATA_PATH, source = "all")
redwood_net_orig <- load_redwood_data(DATA_PATH, source = "net")
redwood_log_orig <- load_redwood_data(DATA_PATH, source = "log")
```

## Quick Look {.tabset .tabset-pills .tabset-square}

### Dates Data

```{r}
dates_orig
skimr::skim(dates_orig)
```

### Motes Data

```{r}
motes_orig
skimr::skim(motes_orig)
```

### Redwood Data {.tabset .tabset-pills .tabset-circle}

#### All data

```{r}
redwood_all_orig
skimr::skim(redwood_all_orig)
```

#### Network data

```{r}
redwood_net_orig
skimr::skim(redwood_net_orig)
```

#### Log data

```{r}
redwood_log_orig
skimr::skim(redwood_log_orig)
```



#### Explorations

There are some NAs in the redwood data. Are they all from the same rows? Let's check.

```{r}
# check for NAs in redwood data
num_nas_all <- sum(apply(redwood_all_orig, 1, function(x) any(is.na(x))))
cat("Number of NAs in all data:", num_nas_all)
num_nas_net <- sum(apply(redwood_net_orig, 1, function(x) any(is.na(x))))
cat("Number of NAs in network data:", num_nas_net)
num_nas_log <- sum(apply(redwood_log_orig, 1, function(x) any(is.na(x))))
cat("Number of NAs in log data:", num_nas_log)
```

These numbers match what we saw in the skim, so let's remember to remove all rows with an NA in the cleaning step.

Is the combined redwood dataset actually a combination of the network and log data? 

```{r}
cat("Number of rows in all redwood data:", nrow(redwood_all_orig))
cat("Number of rows in redwood network data:", nrow(redwood_net_orig))
cat("Number of rows in redwood log data:", nrow(redwood_log_orig))
cat("Number of rows in redwood network + log data:", nrow(redwood_log_orig) + nrow(redwood_net_orig))
```

At least this passes a sanity check. More detailed exploration is needed to confirm further, but we will omit this for brevity.

# Data Cleaning Explorations

Let's look into the weird voltages...

```{r}
merged_voltage_df <- dplyr::left_join(
  redwood_log_orig |> 
    na.omit(),
  redwood_net_orig |> 
    na.omit(),
  by = c("epoch", "nodeid")
)

plt <- merged_voltage_df |> 
  dplyr::filter(
    voltage.x >= 1
  ) |> 
  ggplot2::ggplot() +
  ggplot2::aes(
    x = voltage.x, y = voltage.y
  ) +
  ggplot2::geom_point()
plt
```


Exploring weird temperature and humidity over time...

```{r}
plt_df <- dplyr::bind_rows(
  redwood_log_orig |> 
    na.omit() |> 
    dplyr::mutate(
      source = "log"
    ),
  redwood_net_orig |> 
    na.omit() |> 
    dplyr::mutate(
      source = "network"
    )
)

plt <- plt_df |> 
  dplyr::filter(
    humid_temp >= 0
  ) |> 
  ggplot2::ggplot() +
  ggplot2::aes(
    x = epoch, y = humid_temp, color = source
  ) +
  ggplot2::geom_point()
plt

plt <- plt_df |> 
  dplyr::filter(
    humidity >= 0
  ) |> 
  ggplot2::ggplot() +
  ggplot2::aes(
    x = epoch, y = humidity, color = source
  ) +
  ggplot2::geom_point()
plt
```
