---
title: "Lab 1 - Redwood Data"
author: "Tiffany Tang"
date: "`r Sys.Date()`"
output: vthemes::vmodern
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

for (fname in list.files(here::here("R"), pattern = "*.R")) {
  source(here::here(file.path("R", fname)))
}

DATA_PATH <- here::here("data")
```

# Load Data {.tabset .tabset-vmodern}

```{r load-data}
# load dates data
dates_orig <- load_dates_data(DATA_PATH)

# load motes data
motes_orig <- load_mote_location_data(DATA_PATH)

# load redwood data
redwood_all_orig <- load_redwood_data(DATA_PATH, source = "all")
redwood_net_orig <- load_redwood_data(DATA_PATH, source = "net")
redwood_log_orig <- load_redwood_data(DATA_PATH, source = "log")
```

## Quick Look {.tabset .tabset-pills .tabset-square}

### Dates Data

```{r}
dates_orig
skimr::skim(dates_orig)
```

### Motes Data

```{r}
motes_orig
skimr::skim(motes_orig)
```

### Redwood Data {.tabset .tabset-pills .tabset-circle}

#### All data

```{r}
redwood_all_orig
skimr::skim(redwood_all_orig)
```

#### Network data

```{r}
redwood_net_orig
skimr::skim(redwood_net_orig)
```

#### Log data

```{r}
redwood_log_orig
skimr::skim(redwood_log_orig)
```

#### Explorations

There are some NAs in the redwood data. Are they all from the same rows? Let's check.

```{r}
# check for NAs in redwood data
num_nas_all <- sum(apply(redwood_all_orig, 1, function(x) any(is.na(x))))
cat("Number of NA rows in all data:", num_nas_all)
num_nas_net <- sum(apply(redwood_net_orig, 1, function(x) any(is.na(x))))
cat("Number of NA rows in network data:", num_nas_net)
num_nas_log <- sum(apply(redwood_log_orig, 1, function(x) any(is.na(x))))
cat("Number of NA rows in log data:", num_nas_log)
```

These numbers match what we saw in the skim, so let's remember to remove all rows with an NA in the cleaning step.

Is the combined redwood dataset actually a combination of the network and log data?

```{r}
cat("Number of rows in all redwood data:", nrow(redwood_all_orig))
cat("Number of rows in redwood network data:", nrow(redwood_net_orig))
cat("Number of rows in redwood log data:", nrow(redwood_log_orig))
cat("Number of rows in redwood network + log data:", nrow(redwood_log_orig) + nrow(redwood_net_orig))
```

At least this passes a sanity check. More detailed exploration is needed to confirm further, but we will omit this for brevity.

# Basic Data Cleaning {.tabset .tabset-vmodern}

```{r clean-data}
# load (partially) cleaned dates data
dates_df <- clean_dates_data(dates_orig)

# load (partially) cleaned motes data
motes_df <- clean_mote_location_data(motes_orig)

# load (partially) cleaned redwood data
redwood_all_df <- clean_redwood_data(redwood_all_orig)
redwood_net_df <- clean_redwood_data(redwood_net_orig)
redwood_log_df <- clean_redwood_data(redwood_log_orig)
```

## Quick Look {.tabset .tabset-pills .tabset-square}

### Dates Data

```{r}
dates_df
skimr::skim(dates_df)
```

### Motes Data

```{r}
motes_df
skimr::skim(motes_df)
```

### Redwood Data {.tabset .tabset-pills .tabset-circle}

#### All data

```{r}
redwood_all_df
skimr::skim(redwood_all_df)
```

#### Network data

```{r}
redwood_net_df
skimr::skim(redwood_net_df)
```

#### Log data

```{r}
redwood_log_df
skimr::skim(redwood_log_df)
```

# More Data Cleaning {.tabset .tabset-vmodern}

Let's try to merge all of the relevant data (network/log data, mote locations, and time/dates data) into one nice **clean** data frame.

```{r}
# concatenate log and network data
redwood_df <- dplyr::bind_rows(
  redwood_log_df |>
    dplyr::mutate(source = "log"),
  redwood_net_df |>
    dplyr::mutate(source = "net")
)
```

### Are there duplicates in the data?

```{r}
redwood_distinct <- redwood_df |>
  # To fill in
print(
  sprintf(
    "Number of unique rows in merged log+network: %s (out of %s)",
    nrow(redwood_distinct), nrow(redwood_df)
  )
)
```

Let's remove the duplicates.

```{r}
# To fill in
```

### Merging with dates data

Are all the epochs in redwood data found in the dates data?

```{r}
# To fill in
```

Let's merge the redwood data with the dates data.

```{r}
# To fill in
```

### Merging with mote location data

Are all the motes in the redwood data found in the mote location data?

```{r}
# To fill in
```

Let's merge in the mote location data.

```{r}
# To fill in
```

### Exploring the data

```{r}
redwood_df
```

```{r}
redwood_long_df <- redwood_df |> 
  tidyr::pivot_longer(
    cols = c(voltage, humidity, temp, iPAR, rPAR),
    names_to = "variable",
    values_to = "value"
  )
redwood_long_df
```

```{r}
plt <- redwood_long_df |> 
  ggplot2::ggplot() +
  ggplot2::aes(
    x = datetime, y = value
  ) +
  ggplot2::facet_grid(variable ~ ., scales = "free_y") +
  ggplot2::geom_point()
plt
```

#### What should we do about the 2 trees?

```{r}
plt <- redwood_long_df |> 
  ggplot2::ggplot() +
  ggplot2::aes(
    x = datetime, 
    y = value,
    color = Tree
  ) +
  ggplot2::facet_grid(variable ~ ., scales = "free_y") +
  ggplot2::geom_point()
plt
```

#### What should we do about voltages?

```{r}
merged_voltage_df <- dplyr::inner_join(
  redwood_log_df,
  redwood_net_df,
  by = c("epoch", "nodeid")
)

plt <- merged_voltage_df |> 
  dplyr::filter(
    voltage.x >= 1
  ) |>
  ggplot2::ggplot() +
  ggplot2::aes(
    x = voltage.x, y = voltage.y
  ) +
  ggplot2::geom_point() +
  ggplot2::labs(
    x = "Log Voltage", y = "Network Voltage"
  )
plt
```

```{r}
# fix voltages
```

#### What should we do about outliers/errors?

```{r}
plt <- plot_redwood_over_time(
  data = redwood_df,
  color = "source"
)
plt
```

Getting rid of the obvious errors

```{r}
redwood_df <- redwood_df |> 
  dplyr::filter(
    temp > 0
  )
```

One possible way to remove outliers...

```{r}
thrs <- seq(2.3, 2.5, by = 0.01)
nodes <- unique(redwood_df$nodeid)
epoch_cutoffs <- list()
for (node in nodes) {
  for (thr in rev(thrs)) {
    # find epoch where the node's voltage dips below threshold
    epoch_cutoff <- redwood_df |> 
      dplyr::filter(
        nodeid == !!node,
        voltage < thr
      ) |> 
      dplyr::slice_min(epoch, n = 1)
    if (nrow(epoch_cutoff) == 0) {
      # if voltage does not dip below threshold, do nothing
      next
    }
    redwood_subset <- redwood_df |> 
      dplyr::filter(
        nodeid == !!node,
        epoch >= epoch_cutoff$epoch[[1]],
        epoch <= epoch_cutoff$epoch[[1]] + 100
      ) |> 
      dplyr::arrange(epoch)
    autocor <- as.numeric(acf(redwood_subset$temp, plot = FALSE)[1]$acf)
    if (autocor < 0.8) {
      # if temperature measurements after identified epoch are not highly
        # correlated, then assume node failure and remove data after epoch
      epoch_cutoffs[[as.character(node)]] <- epoch_cutoff$epoch[[1]]
      break
    }
  }
}

redwood_df <- redwood_df |> 
  dplyr::mutate(
    outlier = FALSE
  )
for (node in names(epoch_cutoffs)) {
  redwood_df <- redwood_df |> 
    dplyr::mutate(
      outlier = dplyr::case_when(
        (nodeid == !!node) & (epoch >= epoch_cutoffs[[node]]) ~ T,
        TRUE ~ outlier
      )
    )
}

redwood_df_cleaned <- redwood_df |> 
  dplyr::filter(!outlier)
```

Is this a good outlier removal procedure?

```{r}
# To fill in
```

Looks like we missed some outliers in our initial attempt.

```{r}
# do same for nodes 141 and 145 but use epoch threshold directly since
# voltages are strange
for (node in c(141, 145)) {
  # get all possible epoch cutoffs
  thrs <- redwood_df |> 
    dplyr::filter(
      nodeid == !!node,
      source == "net"
    ) |> 
    dplyr::arrange(epoch) |> 
    dplyr::pull(epoch) |> 
    unique()
  for (thr in thrs) {
    redwood_subset <- redwood_df |> 
      dplyr::filter(
        nodeid == !!node,
        epoch >= !!thr,
        epoch <= !!thr + 100
      ) |> 
      dplyr::arrange(epoch)
    autocor <- as.numeric(acf(redwood_subset$temp, plot = FALSE)[1]$acf)
    if (autocor < 0.8) {
      # if temperature measurements after identified epoch are not highly
      # correlated, then assume node failure and remove data after epoch
      epoch_cutoffs[[as.character(node)]] <- thr
      break
    }
  }
}

redwood_df <- redwood_df |> 
  dplyr::mutate(
    outlier = FALSE
  )
for (node in names(epoch_cutoffs)) {
  # mark outliers
  redwood_df <- redwood_df |> 
    dplyr::mutate(
      outlier = dplyr::case_when(
        (nodeid == !!node) & (epoch >= epoch_cutoffs[[node]]) ~ T,
        TRUE ~ outlier
      )
    )
}

# remove outliers
redwood_df_cleaned <- redwood_df |> 
  dplyr::filter(!outlier)
```

How did we do now?

```{r}
# To fill in
```

# Exploratory Data Analysis

```{r clean-data-continued}
# wrap all cleaning steps into streamlined functions and do merge/cleaning here
redwood_df <- merge_redwood_data(
  dates_data = dates_df,
  motes_data = motes_df,
  redwood_net_data = redwood_net_df,
  redwood_log_data = redwood_log_df
) |> 
  remove_redwood_outliers()
```

What questions can we/do we want to answer with this data?

-   

```{r}

```
