---
title: "Lab 1 - Redwood Data"
author: "Tiffany Tang"
date: "`r Sys.Date()`"
output: vthemes::vmodern
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

for (fname in list.files(here::here("R"), pattern = "*.R")) {
  source(here::here(file.path("R", fname)))
}

DATA_PATH <- here::here("data")
```

# Exploratory Data Analysis {.tabset .tabset-vmodern}

```{r load-data}
# load dates data
dates_df <- load_dates_data(DATA_PATH) |> 
  clean_dates_data()

# load motes data
motes_df <- load_mote_location_data(DATA_PATH) |> 
  clean_mote_location_data()

# load redwood data
redwood_all_df <- load_redwood_data(DATA_PATH, source = "all") |> 
  clean_redwood_data()
redwood_net_df <- load_redwood_data(DATA_PATH, source = "net") |> 
  clean_redwood_data()
redwood_log_df <- load_redwood_data(DATA_PATH, source = "log") |> 
  clean_redwood_data()

# wrap all cleaning steps into streamlined functions and do merge/cleaning here
redwood_df <- merge_redwood_data(
  dates_data = dates_df,
  motes_data = motes_df,
  redwood_net_data = redwood_net_df,
  redwood_log_data = redwood_log_df
) |> 
  remove_redwood_outliers()
```

```{r}
redwood_df
```

What questions can we/do we want to answer with this data?

-   What is the pairwise relationship between variables (temp, humidity, iPAR, rPAR)?

-   Are there differences between the network and the log data?

-   What's the disparity in temperature between the top versus bottom tree?

-   How quickly does the top of the tree warm up?

-   What are the differences in the tree sensor values for different times of day

**What are the pairwise relationships between variables (temp, humidity, iPAR, rPAR)?**

```{r}
plt_df <- redwood_df |> 
  dplyr::slice_sample(n = 50000)
plt <- plt_df |> 
  vdocs::plot_pairs(
    columns = c('temp', "humidity", "iPAR", "rPAR"),
    point_size = 0.1,
    color = plt_df$Height
  )
plt
```

**What's the disparity in temperature between the top versus the bottom of the tree?**

```{r}
plt_df <- redwood_df |> 
  dplyr::group_by(
    Height, time
  ) |> 
  dplyr::summarise(
    mean_temp = mean(temp)
  ) |> 
  dplyr::arrange(Height)
plt <- plt_df |> 
  tidyr::pivot_wider(
    id_cols = Height,
    names_from = time,
    values_from = mean_temp
  ) |> 
  tibble::column_to_rownames("Height") |> 
  vdocs::plot_heatmap()
plt
```

