---
title: "Lab 2 - Linguistics Data"
author: "Tiffany Tang"
date: "`r Sys.Date()`"
output: vthemes::vmodern
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

source(here::here("R", "load.R"))
source(here::here("R", "clean.R"))
source(here::here("R", "plot.R"))

DATA_PATH <- here::here("data")
```

# Load Data {.tabset .tabset-vmodern}

Let's first load in the three datasets:

- `qa_key`: question and answer key
- `ling_orig`: (uncleaned) linguistics survey data
- `zip_df`: mapping between ZIP codes and latitude/longitude coordinates

**Your task:**

- Open `R/load.R` and fill in the code for `load_ling_data()`.

```{r load-data}
# load in data
qa_key <- load_q_and_a_key(DATA_PATH)
ling_orig <- load_ling_data(DATA_PATH)
zip_df <- load_zip_data(DATA_PATH)
```

## Quick Look {.tabset .tabset-pills .tabset-square}

Next, taking a quick look at the data...

### Q and A Data

```{r}
skimr::skim(qa_key)
```

### Linguistics Data

```{r}
skimr::skim(ling_orig)
```

### ZIP Data

```{r}
skimr::skim(zip_df)
```

# Clean Data {.tabset .tabset-vmodern}

In order to facilitate visualizations and future analyses, we will next merge the linguistics survey data with the ZIP locations data and clean the data.

**Your task:**

- Open `R/clean.R` and fill in the code for `merge_ling_zip_data()`. This function should merge the linguistics survey data with the ZIP locations data. The goal is to relate each individual to their latitude/longitude location coordinates.

```{r}
# merge linguistics survey data with ZIP locations
ling_df <- merge_ling_zip_data(ling_orig, zip_df)
skimr::skim(ling_df)
```

**Your task:**

- Open `R/clean.R` and fill in the code for `clean_ling_data()`. Feel free to do any data cleaning you would like here.

```{r}
# do data cleaning (if necessary)
ling_df_cleaned <- clean_ling_data(ling_df)
skimr::skim(ling_df_cleaned)
```

# Data Explorations {.tabset .tabset-vmodern}

Let's do some exploratory data analysis!

To get you started, here is a map of the survey respondents and their responses to question 50.

```{r}
plt <- plot_ling_map(ling_df, qa_key, qid = 50)
plotly::ggplotly(plt)
```

**Your task:**

- Spend some time exploring other questions and visualizations of the data. The goal is to get a feel for what the data is like and what kind of dimension reduction/clustering analyses you might want to do in the future.

```{r}
# TODO: additional explorations
```


# Dimension Reduction {.tabset .tabset-vmodern}

Time to begin dimension reduction.

## Principal Components Analysis {.tabset .tabset-pills .tabset-square}

Let's start with arguably the most popular dimension reduction technique - principal components analysis (PCA).

**Your task:**

- Perform PCA on the linguistics survey data. (Tip: you may want to use the `prcomp()` function from the `stats` package.)

```{r}
# do PCA
```
